---
title: "World Suicide Rate Analysis (1985-2016)"
author: "Wasikul Amin Bipu"
date: "15 November 2023"
execute: 
    cache: true
    echo: true
    output: true
    error: false
    warning: false
code-fold: true
---

## Step by Step World Suicide Rate Analysis with Quarto & R
In this document, I will analyze World Suicide Rates step by step. This will make you understand how I have analyzed the data.

### Data Importing

```{r}
# Importing tidyverse

library("tidyverse")
library("knitr")
library("gridExtra")

# Reading file
db <- read_csv("master.csv")

# Check if the data has imported correctly with glimpse() function
glimpse(db)
```
We can see that all the data has been imported correctly. In this database there are 12 columns and 12 rows. We must check if there is any missing(NA) value in this database. Then we can go to further analysis. Lets check for NA…

### Dimentions and Columns 
I have already imported the data and made it tibble. Now, I should know my data first before going farther. The most basic way to know data is getting its dimention (row and column width) with `dim()` and Knowing column names with `colnames()` function. let's do that practically.

#### Dimention
```{r}
dim(db)
```
Our tibble has 12 columns and 27820 rows. Now lets learn column names

#### Column Names

```{r}
col_dt <- colnames(db)
kable(col_dt, align="l")
```
We have got all the column names. Now we know the basics of our data. Lets go for tidying and wrangling.

### Fixing Variable Types
In tidyverse, when we read external database it assumes most of the data as character type. We will make that factor to analyze them correctly.

```{r}
# Converting variables to factor
db <- db %>% mutate_at(vars("country", "year","sex", "age","generation","country-year"), as.factor)
```

### Renaming Column
Different column are of different type names. I will prefer all the names in same type. I love to prepare column names so that I don’t have to change it for visualization purpose. So, I will allow spaces. Here I am changing column names.

```{r}
db <- db %>% rename("Country"="country","Year"="year","Sex"="sex","Age"="age","No of Suicides"="suicides_no","Population"="population","Suicides (Per 100k Population)"="suicides/100k pop","GDP for Year"="gdp_for_year ($)","GDP per Capita"="gdp_per_capita ($)","Generation"="generation")
```

### Removing Unnecessary Columns
There are few Columns in this database which is really unnecessary. Specially `country-year` column. This column contains data of both year and country which is already available in separate columns. So, I feel that should be removed. Let’s remove the column

```{r}
# removing unnecessary columns
db_2 <- db %>% select(- `country-year`)
```
Now there are only 11 columns. Almost all the data are tidy now. Now go for the further steps. But before performing further steps it would be soothing for eyes to see what data we have now.

### Data Table to Visualize the Dataset
We will use `DT` package to visualize the data in a data table.

```{r}
# importing DT package
library("DT")

# Creating Data Table
datatable(db)
```

### Check NA Values

```{r}
# We will select all the NA value and summarize it
na_val <- db %>% select_if(function(x) any(is.na(x))) %>% summarise_each(funs(sum(is.na(.))))

# Print the data in a table
kable(na_val, align="l")
```

We can see that `HDI for year` contains 19456 NA values. This variable is of type integer. So, for further investigation we will perform univariate analysis of the variable.

### Univariate analysis of `HDI for year` variable
We will create histogram for this analysis. I prefer using ggplot2. You can use any visualization library even vanilla R will do.

```{r}
#| label: fig-hdiyear-hist
#| fig-cap: Histogram of HDI for year
#| fig-align: center


# Generating histogram from HDI for year variable
db %>% ggplot(aes_string("`HDI for year`")) + geom_histogram(bins = 100) + labs(y="Count") + theme_minimal()
```

@fig-hdiyear-hist is generated with 100 bins. In this histogram we are clearly seeing that there are some missing values.

Analysis of data with missing value is bit complicated. So, we will avoid this variable for now. In later part of this document, we will work with this data. To remove the variable we will use `select()` with (-) value.

```{r}
# removing HDI for year column
db_2 <- db %>% select(- `HDI for year`)

# showing the glimpse of the new tibble
glimpse(db_2)
```
Now the database contains only 10 column with no missing values. We will work with this database now.

### Univariate Analysis
I will perform univariate analysis from now. Before performing univariate analysis I must have a plan for what to do and what not to. My plan is following

1. **Summary Statistics:** I will perform summary statistics for all the continuous variables. Summary statistics will include
    i. Mean
    ii. Median
    iii. Inter-Quartile Range
    iv. Standard Deviations
2. **Frequency Table:** for the categorical variables, I will create frequency table to check their conditions.
3. **Charts:** I will generate some plots to visualize the data. such as histogram, density plots, pie charts etc.

Now, let’s start our univariate analysis journey.


### Summary Statistics
We have our plan to summarize the data. We will perform the summary now

```{r}
#| label: tbl-sum-stat
#| tbl-cap: Summary Statistics


# Ensure that the value is in integer not in scientific notation
options(scipen = 999)

# At first select all the variables that are not factor. Then summerise all the data
summary_stat <- db_2 %>% select_if(~ !is.factor(.)) %>% summarise_all(list(mean=mean,sd=sd,min=min,max=max, IQR=IQR, sum=sum), na.rm=TRUE)

# Make Tibble longer to make it easily understandable
summary_stat <- pivot_longer(summary_stat, everything())

# Print the table
kable(summary_stat)
```

I think we have enough data for summary statistics. all the values are in a table. I love it. Let’s go for the next step.

### Frequency Tables
I created summary statistics for all the continuous data. I have some categorical data also. I have to check them using frequency table. So, without wasting time i will go for frequency tables

```{r}
#| label: tbl-freq-country
#| tbl-cap: Frequency Table of Country


tab_country <- table(db_2$Country)

kable(tab_country)
```

```{r}
#| label: tbl-freq-year
#| tbl-cap: Frequency Table of Year

tab_year <- table(db_2$Year)

kable(tab_year)
```

```{r}
#| label: tbl-freq-sex
#| tbl-cap: Frequency Table of Sex


tab_sex <- table(db_2$Sex)

kable(tab_sex)
```

```{r}
#| label: tbl-freq-age
#| tbl-cap: Frequency Table of Age


tab_age <- table(db_2$Age)

kable(tab_age)
```

```{r}
#| label: tbl-freq-gen
#| tbl-cap: Frequency Table of Generation


tab_gen <- table(db_2$Generation)

kable(tab_gen)
```

### Charts
In this section of univariate analysis we will visualize all the data in different types of plots. I wish the plots will amaze you. Let’s do it.

1. **Country Data Analysis:** For Analysis of country data we will create several plots. I don’t think visualizing all the country data will be appropriate. So, I will take sample of 5 data from the frequency table to visualize.

```{r}
#| label: fig-bar-country
#| fig-cap: Bar Chart of Frequency of Random Countries


dat <- as.data.frame(tab_country) 

sample_n(dat,5) %>% ggplot(aes(Var1,Freq, fill=Var1)) + geom_col() + labs(x="Countries",y="Frequency") + guides(fill=FALSE)
```

2. **Year:** I don’t think this analysis add any values. So, i will avoid this variable.
3. **Sex:** All the variables here has same frequency. So, we can not visualize the data
4. **Age:** Here there are also same frequency. So we will avoid this.
5. ***Generation:*** Generation Data have different frequency value. We will create a lollypop plot with this.

```{r}
#| label: fig-lolly-gen
#| fig-cap: Lollypop Plot of Frequency of Generation

dat <- as.data.frame(tab_gen) 

dat %>% ggplot(aes(Var1,Freq)) + geom_point(size=5) + geom_segment(aes(x=Var1,xend=Var1, y=0,yend=Freq)) + labs(x="Generation",y="Frequency")
```
### For Continuous Variables

### Histogram, Density Plot and Frequency Polygon
I will create Histogram, Density Plot and Frequency Polygon of several continuous variables for the univariable analysis. These plots will help us understand about the variables. Let's start our analysis.

#### No of Suicides

```{r}
#| label: fig-no-sui
#| fig-cap: Histogram, Density plot and Frequency Polygon of No of Suicides

plot_no_sui <- db %>% ggplot(aes(`No of Suicides`))

hist_no_sui <- plot_no_sui + geom_histogram(width=0.5) + labs(title="Histogram of No of Suicides", y="Count")


den_no_sui <- plot_no_sui + geom_density() + labs(title="Density plot of No of Suicides",y="Density")

freqpol_no_sui <- plot_no_sui + geom_freqpoly() + labs(title="Frequency Polygon of No of Suicides",y="Count")

grid.arrange(hist_no_sui, den_no_sui, freqpol_no_sui, ncol=2)
```
#### Population


```{r}
#| label: fig-pop
#| fig-cap: Histogram, Density plot and Frequency Polygon of Population

plot_pop <- db %>% ggplot(aes(Population))
hist_pop <- plot_pop + geom_histogram(width=0.5) + labs(title="Histogram of Population", y="Count")


den_pop <- plot_pop + geom_density() + labs(title="Density plot of Population",y="Density")

freqpol_pop <- plot_pop + geom_freqpoly() + labs(title="Frequency Polygon of Population", y="Count")

grid.arrange(hist_pop, den_pop, freqpol_pop, ncol=2)
```


#### Suicides (Per 100k Population) 

```{r}
#| label: fig-sui
#| fig-cap: Histogram, Density plot and Frequency Polygon of Suicides (Per 100k Population) 

plot_sui <- db %>% ggplot(aes(`Suicides (Per 100k Population)`))

hist_sui <- plot_sui + geom_histogram(width=0.5) + labs(title="Histogram of Suicides\n(Per 100k Population)", y="Count")


den_sui <- plot_sui + geom_density() + labs(title="Density plot of Suicide\n(Per 100k Population)",y="Density")

freqpoly_sui <- plot_sui + geom_freqpoly() + labs(title="Frequency Polygon of Suicide\n(Per 100k Population) ", y="Count")

grid.arrange(hist_sui, den_sui, freqpoly_sui, ncol=2)
```

#### GDP for Year 

```{r}
#| label: fig-gdp
#| fig-cap: Histogram, Density plot and Frequency Polygon of GDP for Year 

plot_gdp <- db %>% ggplot(aes(`GDP for Year`))

hist_gdp <- plot_gdp + geom_histogram(width=0.5) + labs(title="Histogram of GDP for Year", y="Count")


den_gdp <- plot_gdp+ geom_density() + labs(title="Density plot of GDP for Year",y="Density")

freqpoly_gdp <- plot_gdp + geom_freqpoly(width=0.5) + labs(title="Frequency Polygon of GDP for Year", y="Count")

grid.arrange(hist_gdp, den_gdp, freqpoly_gdp, ncol=2)
```

#### GDP per Capita


```{r}
#| label: fig-gdp-capita
#| fig-cap: Histogram, Density plot and Frequency Polygon of GDP for Year 
plot_gdp_capita <- db %>% ggplot(aes(`GDP per Capita`))

hist_gdp_capita <- plot_gdp_capita + geom_histogram(width=0.5) + labs(title="Histogram of GDP per Capita", y="Count")


den_gdp_capita <- plot_gdp_capita + geom_density() + labs(title="Density plot of GDP per Capita",y="Density")

freqpol_gdp_capita <- plot_gdp_capita + geom_freqpoly() + labs(title="Frequency Polygon of GDP per Capita")

grid.arrange(hist_gdp_capita, den_gdp_capita, freqpol_gdp_capita, ncol=2)
```

### Box Plot and Violin Plot
Although we will use multiple variables to create box-whisker plot or Violin Plot. it is considered as univariate analysis. we will create as much boxplot and violin plot as possible for the analysis.

As we have suicide data and most of the value is around zero but there are a higher values too. So boxplot or violin plot is not the best way to visualize it.The plot will look strange for this analysis. So, I will not create many. Though I will create some to ensure that all types of visualization is present in this project.

```{r}
#| label: fig-box-no-sui
#| fig-cap: Box Plot and Violin Plot of No of suicides

p1 <- db %>% ggplot(aes(Sex,`Suicides (Per 100k Population)`)) + ylim(-100,300) + geom_boxplot() + labs(title="Box Plot of Suicides (Per 100k population) in different Gender", x="Gender")

p2 <- db %>% ggplot(aes(Sex,`Suicides (Per 100k Population)`, fill=Sex)) + ylim(-100,300) + geom_violin() + labs(title="Violin Plot of Suicides (Per 100k population) in different Gender", x="Gender") + guides(fill=FALSE)

grid.arrange(p1,p2,ncol=1)
```


```{r}
#| label: fig-box-sui-gen
#| fig-cap: Box Plot and Violin Plot of No of suicides

p1 <- db %>% ggplot(aes(Generation,`Suicides (Per 100k Population)`)) + ylim(-100,300) + geom_boxplot() + labs(title="Box Plot of Suicides (Per 100k population) in different Generations") 

p2 <- db %>% ggplot(aes(Generation,`Suicides (Per 100k Population)`,fill=Generation)) + ylim(-100,300) + geom_violin() + labs(title="Violine Plot of Suicides (Per 100k population) in different Generations") +guides(fill=FALSE)

grid.arrange(p1,p2,ncol=1)
```

## Bivariate Analysis
In this section we will discover relationship between two variables. We will do step by step analysis.



## Multivariate Analysis
In this section we will discover relationship among multiple vaiables. For this section, I will do some basic tests and generate some basic plots. Then we will develop some research questions to analyze the data perfectly. Then according to the research questions we will create some plots and perform some tests to find out the answers. I hope you will be with me in the whole journey.

### References
1. The Data is downloaded from the following site [Data Frame](https://www.kaggle.com/datasets/russellyates88/suicide-rates-overview-1985-to-2016/data)